---
title: 'Session 1: Transformations into stationarity'
Date: 08/02/2018
output:
  word_document: default
  pdf_document: default
  html_document: default
Authors: JAS + LA-extension
---

### RMarkdown Introduction

[https://rmarkdown.rstudio.com/](RMarkdown)

[https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-spanish.pdf]()

[https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf]()

### Nottem

Average Monthly Temperatures at Nottingham, 1920-1939

Description: A time series object containing average air temperatures at Nottingham Castle in degrees Fahrenheit for 20 years.

```{r}
#?nottem
nottem # List Nottem observations
```
```{r}
##LA
# Some useful properties of a time series object
class(nottem) #Type of object
str(nottem) #Type  plus atributes

length(nottem)         #No. of observations
(stno=start(nottem))   #Starting point ...
(enno=end(nottem))     #Ending point ...
frequency(nottem)      # No. of (months) in a (year)

(enno[1]-stno[1]+1)
(enno[1]-stno[1]+1) * frequency(nottem) #No. of observations
```
####Is Variance constant?
```{r}
plot(nottem,type="o")
#LA
abline(h=mean(nottem), col=2)
#boxplot(nottem~floor(time(nottem))) #boxplot x cada año
# m=apply(matrix(nottem,ncol=20),2,mean) #lleno matrix 12x20 x columnas y get media por cada año
# v=apply(matrix(nottem,ncol=20),2,var)
# plot(v~m,main="Mean-Variance plot")
# abline(lm(v~m))
```
#####The varianza seemingly constant!  

####Is seasonality present? 
```{r}
#LA
plot(window(nottem,start=1920,end=1925,type="o"))   #Zoom
```
```{r}
monthplot(nottem)
```
```{r}
# ACF decreasing slowly => non-stationarity
acf(nottem,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)),lwd=2)
```

#####Yes, a pattern repeats itself every s=12 months
```{r}
##
# Difference of order 12 to remove the seasonal component 
d12nottem=diff(nottem,12)
```

####Is the mean constant? 
```{r}
plot(d12nottem)
abline(h=0)
```
```{r}
#var(d12nottem)
#var(diff(d12nottem))
```

#####Yes, time series seemingly constant with mean around 0
####Is the transformed series already stationary?
```{r}
# Constant mean, constant variance, ACF with fast decay => stationarity
par(mfrow=c(1,2))
acf(d12nottem,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)),lwd=2)
pacf(d12nottem,ylim=c(-1,1),lag.max=60,col=c(rep(1,11),2),lwd=2)
par(mfrow=c(1,1))

#monthplot(d12nottem)
```

#####Summary: Series d12Nottem has constant mean and variance.ACF decaying rapidly towards 0. => Stationarity achieved





### GNPSH
Gross National Product USA, Quarterly data (I/1947-I/1991) Seasonally adjusted

Description: Gross national product (GNP) is the market value of all the products and services produced in one year by labor and property supplied by the citizens of a country[https://en.wikipedia.org/wiki/Gross_national_product](wiki).

```{r}
# Object ts (time series): starting in 1947 and frequency 4 (quarterly data)
(gnpsh=ts(read.table("gnpsh.dat")[,1],start=1947,freq=4))
plot(gnpsh)
#LA
start(gnpsh)        #Starting point ...
end(gnpsh)          #Ending point ...
frequency(gnpsh)    # No. of (quarters) in a (year)
length(gnpsh)       #No. of observations 
#(end(gnpsh)[1]-start(gnpsh)[1]+1)*4 -3
```
####Is Variance constant?
```{r}
w1 <- window(gnpsh,start=1947,end=1960)
w2 <- window(gnpsh,start=1960,end=1970)
w3 <- window(gnpsh,start=1970,end=1990)

par(mfrow=c(1,3))
plot(w1,ylim=c(min(gnpsh),max(gnpsh)))
plot(w2,ylim=c(min(gnpsh),max(gnpsh)))
plot(w3,ylim=c(min(gnpsh),max(gnpsh)))

par(mfrow=c(1,3))
boxplot(w1,ylim = c(min(gnpsh),max(gnpsh)))
boxplot(w2,ylim = c(min(gnpsh),max(gnpsh)))
boxplot(w3,ylim = c(min(gnpsh),max(gnpsh)))
```

##### Variance appears to be non-constant: it seems to change with the level of the series
```{r}
# Logarithm transformation to stabilize the variance
lngnpsh=log(gnpsh)
plot(lngnpsh)
```
```{r}
lnw1 <- window(lngnpsh,start=1947,end=1960)
lnw2 <- window(lngnpsh,start=1960,end=1970)
lnw3 <- window(lngnpsh,start=1970,end=1990)

par(mfrow=c(1,3))
plot(lnw1,ylim=c(min(lngnpsh),max(lngnpsh)))
plot(lnw2,ylim=c(min(lngnpsh),max(lngnpsh)))
plot(lnw3,ylim=c(min(lngnpsh),max(lngnpsh)))

par(mfrow=c(1,3))
boxplot(lnw1,ylim=c(min(lngnpsh),max(lngnpsh)))
boxplot(lnw2,ylim=c(min(lngnpsh),max(lngnpsh)))
boxplot(lnw3,ylim=c(min(lngnpsh),max(lngnpsh)))
```

#####Now the variance constant...

####Is seasonality present? Nop, GNPSH series has been Seasonally adjusted

####Is the mean constant? Cleary nop
```{r}
par(mfrow=c(1,2))
acf(lngnpsh,ylim=c(-1,1),lag.max=60,lwd=2)
pacf(lngnpsh,ylim=c(-1,1),lag.max=60,lwd=2)
par(mfrow=c(1,1))
```
#####Mean clearly non-constant because of a global linear trend => Non stationarity
```{r}
# Regular difference to remove global linear trend
d1lngnpsh=diff(lngnpsh) #LA: show difln equiv tasaincremrelativo ln(xt/xt-1)=(xt-xt-1)/xt-1
plot(d1lngnpsh)
abline(h=0)
```
```{r}
# Now we have constant mean (different from zero!), constant variance and ACF with few significants lags => Stationarity
par(mfrow=c(1,2))
acf(d1lngnpsh,ylim=c(-1,1),lag.max=60,lwd=2)
pacf(d1lngnpsh,ylim=c(-1,1),lag.max=60,lwd=2)
par(mfrow=c(1,1))
```
#####Summary: Series d1lngnpsh has constant mean and variance.ACF decaying rapidly towards 0. => Stationarity achieved



### AirPassengers
Monthly Airline Passenger Numbers 1949-1960

Description:  The classic Box & Jenkins airline data. Monthly totals of international airline passengers, 1949 to 1960.
```{r}
AirPassengers
plot(AirPassengers,type="o")

#LA: 
(start(AirPassengers))
(end(AirPassengers))
(fq=frequency(AirPassengers))
length(AirPassengers)
```

```{r}
boxplot(AirPassengers~floor(time(AirPassengers)))

m=apply(matrix(AirPassengers,ncol=12),2,mean)
v=apply(matrix(AirPassengers,ncol=12),2,var)
plot(v~m,main="Mean-Variance plot")

```

##### Variance is not constant: higher when the level raises => Non-Stationarity
```{r}
# Logarithms of the data to stabilize the variance
lnAirPassengers=log(AirPassengers)
plot(lnAirPassengers,type="o")

boxplot(lnAirPassengers~floor(time(lnAirPassengers)))

m=apply(matrix(lnAirPassengers,ncol=12),2,mean)
v=apply(matrix(lnAirPassengers,ncol=12),2,var)
plot(v~m,main="Mean-Variance plot")

```

##### Now, the variance can be considered constant => Solved
```{r}
# Similar pattern each year (12 observations) implies seasonality => Non-Sationarity
monthplot(lnAirPassengers)

plot(decompose(lnAirPassengers))
```
```{r}
# Difference of order 12 to remove the seasonal component
d12lnAirPassengers<-diff(lnAirPassengers,lag=12)
plot(d12lnAirPassengers,main="d12lnAirPassengers")
abline(h=0)
```

#####Seasonality removed!
```{r}
# Constant mean? Not clear! 
# Another Regular Difference to remove unit roots

d1d12lnAirPassengers<-diff(d12lnAirPassengers)
plot(d1d12lnAirPassengers,main="d1d12lnAirPassengers")
abline(h=0)
```
```{r}
# Constant mean, constant variance (!) and fast decay in the ACF =>Stationarity
par(mfrow=c(1,2))
acf(d1d12lnAirPassengers,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)),lwd=2)
pacf(d1d12lnAirPassengers,ylim=c(-1,1),lag.max=60,col=c(rep(1,11),2),lwd=2)
par(mfrow=c(1,1))
```
```{r}
# Let's check the variances to decide the number of differences
var(lnAirPassengers)
var(d12lnAirPassengers)
var(d1d12lnAirPassengers)
var(diff(d1d12lnAirPassengers))

```
#####Summary: Series d1d12lnAirPassengers has constant mean and variance.ACF decaying rapidly towards 0. => Stationarity achieved


### RANDOM
200 observations of White Noise with variance=1 (independent!)

```{r}
# By fixing the seed of simulation, we can reproduce the pseudo-random sequence
set.seed(123456)
RANDOM=rnorm(200, mean=0, sd=1)
ts.plot(RANDOM)
abline(h=0)

# Constant mean and variance are clear

# Normal distribution verified:

# Histogram with theoretical distribution overlaped
hist(RANDOM,prob=T)
curve(dnorm,col=2,add=T)

# Normal-probability plot (quantile-qunatile)
qqnorm(RANDOM)
qqline(RANDOM,lwd=2,col=2)

# Shapiro-Wilks Test for normality
shapiro.test(RANDOM)


# Independence verified:

# All lags in the ACF/PACF are non-significant
par(mfrow=c(1,2))
acf(RANDOM,ylim=c(-1,1),lag.max=60,lwd=2)
pacf(RANDOM,ylim=c(-1,1),lag.max=60,lwd=2)
par(mfrow=c(1,1))

# Durbin-Watson test for autocorrelation
library(lmtest)
dwtest(RANDOM~1)
```



#### SUNSPOT
Yearly Sunspot Data, 1700-1988

```{r}
sunspot.year
plot(sunspot.year,type="o")

# It seems a similar pattern repeating, but the frequency is not constant!
# It's not seasonality but cycles!!!
# Difference is not a solution. AR(p) with large p is!
par(mfrow=c(1,2))
acf(sunspot.year,ylim=c(-1,1),lag.max=60,lwd=2)
pacf(sunspot.year,ylim=c(-1,1),lag.max=60,lwd=2)
par(mfrow=c(1,1))
```

